---
title: "Timely and Effective Care Across U.S. States"
subtitle: "Crowding drives ER waits · Care quality is not one single trait · INFO 526 Final Project"
author: "Christopher Paskins"
format:
  revealjs:
    theme: [default, data/customtheming.scss]
    slide-number: c/t
    show-slide-number: all
    transition: fade
    transition-speed: fast
    incremental: true
    highlight-style: github
    fig-align: center
editor: visual
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(maps)

care <- read_csv("data/care_state.csv")
pop  <- read_csv("data/state_population.csv")

# Figures styled to blend into the dark slides
theme_slides <- theme_minimal(base_size = 15) +
  theme(
    plot.background  = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA),
    legend.background = element_rect(fill = "transparent", color = NA),
    panel.grid.major = element_line(color = "#2b4258", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    text        = element_text(color = "#e2e8f0"),
    axis.text   = element_text(color = "#b6c2cf"),
    axis.title  = element_text(color = "#e2e8f0"),
    plot.title  = element_text(color = "#ffffff", face = "bold"),
    strip.text  = element_text(color = "#e2e8f0", face = "bold"),
    legend.text = element_text(color = "#e2e8f0"),
    legend.title = element_text(color = "#e2e8f0")
  )
theme_set(theme_slides)
knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

# One consistent region color mapping used across every categorical chart
region_pal <- c("Northeast" = "#2dd4bf", "Midwest" = "#fbbf24",
                "South" = "#f472b6", "West" = "#60a5fa")

region_lookup <- tibble(state = state.abb, region = as.character(state.region)) |>
  mutate(region = recode(region, "North Central" = "Midwest")) |>
  bind_rows(tibble(state = "DC", region = "South")) |>
  mutate(region = factor(region, levels = c("Northeast", "Midwest", "South", "West")))

ed_overall <- care |>
  filter(measure_id == "OP_18b") |>
  select(state, wait_min = score) |>
  inner_join(region_lookup, by = "state") |>
  inner_join(pop, by = "state")

ed_tiers <- care |>
  filter(measure_id %in% c("OP_18b_LOW_MIN", "OP_18b_MEDIUM_MIN",
                           "OP_18b_HIGH_MIN", "OP_18b_VERY_HIGH_MIN")) |>
  transmute(state, score,
            tier = recode(measure_id,
                          OP_18b_LOW_MIN = "Low", OP_18b_MEDIUM_MIN = "Medium",
                          OP_18b_HIGH_MIN = "High", OP_18b_VERY_HIGH_MIN = "Very High")) |>
  mutate(tier = factor(tier, levels = c("Low", "Medium", "High", "Very High"))) |>
  inner_join(region_lookup, by = "state") |>
  drop_na(score)

# All three effectiveness measures re-oriented so HIGHER = BETTER
effect <- care |>
  filter(measure_id %in% c("OP_23", "OP_29", "SAFE_USE_OF_OPIOIDS")) |>
  select(state, measure_id, score) |>
  pivot_wider(names_from = measure_id, values_from = score) |>
  mutate(avoid_opioids = 100 - SAFE_USE_OF_OPIOIDS) |>
  inner_join(region_lookup, by = "state")

effect_long <- effect |>
  pivot_longer(c(OP_23, OP_29, avoid_opioids),
               names_to = "measure", values_to = "score") |>
  mutate(measure = recode(measure,
    OP_23 = "Stroke scan in 45 min",
    OP_29 = "Colonoscopy follow-up",
    avoid_opioids = "Avoided concurrent opioids")) |>
  drop_na(score)

pairs <- bind_rows(
  effect |> transmute(pair = "Stroke vs colonoscopy", x = OP_23, y = OP_29),
  effect |> transmute(pair = "Stroke vs opioid safety", x = OP_23, y = avoid_opioids),
  effect |> transmute(pair = "Colonoscopy vs opioid safety", x = OP_29, y = avoid_opioids)
) |>
  drop_na()

pair_r <- pairs |>
  group_by(pair) |>
  summarise(r = cor(x, y), xpos = min(x), ypos = max(y), .groups = "drop")

# Choropleth frame for ER wait time
wait_map <- tibble(state = state.abb, mapname = tolower(state.name)) |>
  inner_join(ed_overall |> select(state, wait_min), by = "state")
map_df <- map_data("state") |>
  left_join(wait_map, by = c("region" = "mapname"))

# Inline numbers
r_pop   <- round(cor(log10(ed_overall$population), ed_overall$wait_min), 2)
ne_wait <- round(median(ed_overall$wait_min[ed_overall$region == "Northeast"]))
mw_wait <- round(median(ed_overall$wait_min[ed_overall$region == "Midwest"]))
low_w   <- round(median(ed_tiers$score[ed_tiers$tier == "Low"]))
vhigh_w <- round(median(ed_tiers$score[ed_tiers$tier == "Very High"]))
gap_min <- vhigh_w - low_w
n_q2    <- nrow(drop_na(effect, OP_23, OP_29, avoid_opioids))
r_2329  <- round(cor(effect$OP_23, effect$OP_29, use = "complete.obs"), 2)
r_23op  <- round(cor(effect$OP_23, effect$avoid_opioids, use = "complete.obs"), 2)
r_29op  <- round(cor(effect$OP_29, effect$avoid_opioids, use = "complete.obs"), 2)
```

## Bottom line up front {.center}

[The two findings]{.kicker}

-   **ER waits are about crowding, not geography or size.** How busy a room is drives waiting far more than which state it is in, and big states are not the slow ones.
-   **Care quality is not one single trait.** A state strong on one kind of care is no more likely to be strong on the others. This was the surprise.

## Why this matters

-   Long emergency room waits and uneven care quality are the most visible signs of a health system under strain.
-   The federal government measures both, state by state, but the numbers rarely get told as a story.
-   I set out to answer two plain questions and let the data push back on my assumptions.

## The data

::: bigstat
`r n_distinct(care$measure_id)` measures · `r n_distinct(care$state)` jurisdictions · 2 years
:::

[CMS "Timely and Effective Care," via TidyTuesday 2025-04-08]{.statlabel}

-   One row per state and measure, reshaped for each question.
-   I added two variables of my own: each state's **Census region** and its **2024 population**.
-   All findings are **associations**, not causes, since the data are observational.

# Question 1 {background-color="#0d6b73"}

How long do patients wait in the ER, and why?

## The Northeast waits the longest

::::: columns
::: {.column width="62%"}
```{r}
#| label: q1-map
#| fig-width: 6.6
#| fig-asp: 0.66
ggplot(map_df, aes(long, lat, group = group, fill = wait_min)) +
  geom_polygon(color = "#0b1622", linewidth = 0.15) +
  coord_quickmap() +
  scale_fill_gradient(low = "#a7f3d0", high = "#0f766e",
                      na.value = "#334155", name = "Minutes") +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "transparent", color = NA),
    legend.text  = element_text(color = "#e2e8f0"),
    legend.title = element_text(color = "#e2e8f0")
  )
```
:::

::: {.column width="38%"}
-   Darker states wait longer.
-   The **Northeast** is slowest, at a median of **`r ne_wait` min**.
-   The **Midwest** is fastest, at **`r mw_wait` min**.
-   Region matters, but as we will see, it is not the main driver.
:::
:::::

## Busier rooms, longer waits

::::: columns
::: {.column width="60%"}
```{r}
#| label: q1-tiers
#| fig-width: 6.4
#| fig-asp: 0.82
slope_df  <- ed_tiers |>
  group_by(region, tier) |>
  summarise(med = median(score), .groups = "drop")
slope_lab <- filter(slope_df, tier == "Very High")

ggplot(slope_df, aes(x = tier, y = med, color = region, group = region)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.8) +
  geom_text(data = slope_lab, aes(label = region),
            hjust = -0.12, size = 4.2, fontface = "bold") +
  scale_color_manual(values = region_pal) +
  scale_x_discrete(expand = expansion(add = c(0.15, 1.2))) +
  labs(x = "ER volume tier", y = "Median time in ER (minutes)") +
  theme(legend.position = "none")
```
:::

::: {.column width="40%"}
-   Wait climbs from about **`r low_w` min** to **`r vhigh_w` min** as volume rises.
-   That is **`r gap_min` extra minutes**, more than an hour longer.
-   **Every region** shows the same upward slope.
:::
:::::

## Size is not the culprit

::::: columns
::: {.column width="60%"}
```{r}
#| label: q1-pop
#| fig-width: 6.2
#| fig-asp: 0.9
ggplot(ed_overall, aes(x = population, y = wait_min)) +
  geom_point(aes(color = region), size = 3, alpha = 0.9) +
  geom_smooth(method = "lm", se = FALSE, color = "#fbbf24", linetype = "dashed") +
  annotate("text", x = 7e5, y = max(ed_overall$wait_min),
           label = paste0("r = ", r_pop, "  (essentially flat)"),
           color = "#fbbf24", hjust = 0, vjust = 1, size = 5, fontface = "bold") +
  scale_x_log10(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  scale_color_manual(values = region_pal) +
  labs(x = "State population (log scale)",
       y = "Median time in ER (minutes)", color = "Region")
```
:::

::: {.column width="40%"}
-   Population barely moves wait time (correlation of only **`r r_pop`** on a log scale).
-   Big states are **not** the slow ones.
-   Crowding, not size, is what counts.
:::
:::::

# Question 2 {background-color="#0d6b73"}

If a state does well on one kind of care, does it do well on the others?

## Regional gaps are small

::::: columns
::: {.column width="58%"}
```{r}
#| label: q2-region
#| fig-width: 6.4
#| fig-asp: 0.95
effect_long |>
  group_by(region, measure) |>
  summarise(mean_score = mean(score), .groups = "drop") |>
  ggplot(aes(x = region, y = mean_score, fill = region)) +
  geom_col(show.legend = FALSE, alpha = 0.95) +
  geom_text(aes(label = round(mean_score, 1)), vjust = -0.4,
            color = "#e2e8f0", size = 4.2) +
  facet_wrap(~ measure, ncol = 1, scales = "free_y") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  scale_fill_manual(values = region_pal) +
  labs(x = NULL, y = "Regional average (percent)")
```
:::

::: {.column width="42%"}
-   Three very different measures, all oriented so **higher is better**.
-   No region pulls clearly ahead on **all three** at once.
-   The differences between regions are **small**.
:::
:::::

## No all-around winners

::::: columns
::: {.column width="48%"}
```{r}
#| label: q2-pairs
#| fig-width: 5.0
#| fig-asp: 1.15
ggplot(pairs, aes(x = x, y = y)) +
  geom_point(alpha = 0.8, color = "#2dd4bf", size = 2.2) +
  geom_smooth(method = "lm", se = FALSE, color = "#fbbf24", linetype = "dashed") +
  geom_text(data = pair_r, aes(x = xpos, y = ypos, label = paste0("r = ", sprintf("%+.2f", r))),
            color = "#fbbf24", hjust = 0, vjust = 1, size = 4.4,
            fontface = "bold", inherit.aes = FALSE) +
  facet_wrap(~ pair, ncol = 1, scales = "free") +
  labs(x = "First measure (percent)", y = "Second measure (percent)")
```
:::

::: {.column width="52%"}
-   Every pairing is a loose cloud with a nearly flat trend.
-   Correlations near zero: **`r r_2329`**, **`r r_23op`**, **`r r_29op`**.
-   **None are statistically significant** (n = `r n_q2`).
-   Doing well on one measure **does not predict** the others.
:::
:::::

## Conclusions {.center}

[What it means]{.kicker}

-   **Crowding drives ER waits far more than geography or state size**, and big states are not the slow ones.
-   **State care quality is not one single trait.** Strength in one area travels only weakly to others.
-   These are associations from one snapshot, so they describe patterns, not causes.
-   **Recommendations and next steps:** add a rural versus urban dimension, if the data can be found, to drill further into where these differences come from.

## Thank you {.center}

[Christopher Paskins · INFO 526]{.kicker}

-   Data: CMS Timely and Effective Care, via TidyTuesday 2025-04-08.
-   Full write-up and code: project website and repository.
