---
title: "Timely and Effective Care Across U.S. States"
subtitle: "Proposal"
author:
  - name: "Christopher Paskins"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "A state-level look at emergency-department throughput and treatment-effectiveness using CMS Timely and Effective Care data."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
  pdf:
    toc: false
    code-overflow: wrap
    geometry: margin=1in
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false
care <- read_csv("data/care_state.csv")
state_pop <- read_csv("data/state_population.csv")
glimpse(care)
```

The dataset is the **Centers for Medicare & Medicaid Services (CMS) "Timely and Effective Care"** collection, sourced from [data.cms.gov](https://data.cms.gov) and curated for [TidyTuesday week 14, 2025](https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-04-08). It reports state-level hospital quality measures for reporting periods spanning 2023–2024. The file `care_state.csv` holds `r nrow(care)` observations of `r ncol(care)` variables, covering `r n_distinct(care$state)` jurisdictions (50 states, DC, and U.S. territories) and `r n_distinct(care$measure_id)` distinct measures grouped into `r n_distinct(care$condition)` condition areas. The data are stored in a long/tidy layout, one row per state-and-measure, so each analysis pivots the relevant subset wider. The key variables are `state` and `condition` (categorical), `measure_id`/`measure_name` (the metric being reported), and `score` (the numerical value, whose units and direction differ by measure).

I chose this dataset because emergency-department crowding and hospital care quality are active national policy concerns, the data are recent and federally maintained, and the structure cleanly satisfies our need for both numerical (`score`, state population) and categorical (`condition`, Census `region`) variables. It also lets us recode a meaningful geographic grouping and merge in external population data to enrich the analysis.

I derive two variables used across the questions. **`region`** is the four-level U.S. Census region (Northeast, Midwest, South, West) recoded from each state code, and **`state_pop`** is the state's 2024 population, merged from a small Census-estimate file committed to `/data`.

## Questions

**Question 1 — Emergency-department throughput.** How does the median time patients spend in the emergency department before leaving (`OP_18b`) vary by **ED volume tier** (low through very high, captured by the `OP_18b_*_MIN` measures) and **Census region**, and is it associated with **state population**?

**Question 2 — Treatment effectiveness.** Across **Census regions**, how do states compare on **treatment-effectiveness** measures, stroke CT timeliness (`OP_23`), colonoscopy follow-up appropriateness (`OP_29`), and safe opioid use (`SAFE_USE_OF_OPIOIDS`), and do states that score well on one tend to score well on the others?

These questions are distinct: Question 1 uses the ED-timing measures plus population, while Question 2 uses three effectiveness measures; they share only the derived `region` variable. Because the data are observational, we will frame all findings as **associations** and avoid causal language.

## Analysis plan

**Question 1 (ED throughput).** Filter to `condition == "Emergency Department"`, keep the volume-stratified `OP_18b_LOW_MIN`, `OP_18b_MEDIUM_MIN`, `OP_18b_HIGH_MIN`, and `OP_18b_VERY_HIGH_MIN` measures, and pivot so each state has one wait-time value per volume tier. Join `region` and `state_pop`. Variables involved: ED wait time (numerical), volume tier (categorical), region (categorical), and state population (numerical), four variables. Planned `ggplot2` figures: (1) boxplots of wait time by volume tier, colored or faceted by region; (2) a scatter of state population versus median ED wait time.

**Question 2 (treatment effectiveness).** Filter to the three effectiveness measures, pivot them wide so each state has one column per measure, and join `region`. We will first confirm each measure's direction ("higher is better" versus "lower is better") from `measure_name`. Variables involved: `OP_23`, `OP_29`, and `SAFE_USE_OF_OPIOIDS` scores (numerical) plus region (categorical), more than two variables. Planned `ggplot2` figures: (1) a regional comparison of the measures (e.g., point-range or bar summaries by region); (2) a scatter relating two measures across states to assess whether strong performance on one accompanies strong performance on another.

**Derived variables.** `region` is built from base R's `state.abb`/`state.region` (recoding "North Central" to "Midwest" and assigning DC to the South). `state_pop` is merged from `data/state_population.csv`. Territories lacking a Census region or population estimate will be dropped from the relevant analyses and this exclusion will be reported.

## Weekly plan of attack

| Task Name | Status | Due | Priority | Summary |
|---------------|---------------|---------------|---------------|---------------|
| Finalize dataset & draft proposal | Complete | Jun 4 | High | Load data, write Q1/Q2, complete proposal.qmd and data README |
| Submit proposal for peer review | Complete | Jun 4, 5pm | High | Render and push proposal |
| Address peer-review issues | Not started | Jun 9, 5pm | High | Revise proposal and close GitHub issues |
| Data wrangling (region, population, pivots) | Not started | Jun 13 | High | Build clean tidy frames for both questions |
| Question 1 analysis & figures | Not started | Jun 18 | Moderate | ED throughput plots and discussion |
| Question 2 analysis & figures | Not started | Jun 23 | Moderate | Effectiveness plots and discussion |
| Write-up (index.qmd) & website | Not started | Jun 27 | Moderate | Introduction, both questions, conclusions |
| Slides & recorded presentation | Not started | Jun 30, 5pm | High | \~6 content slides, presentation under 5 minutes |